---
title: "Univariate Graphing Assignment"
author: YOUR NAME HERE
format: pdf
execute:
  echo: true
  warning: false
  message: false
---


```{r}
library(tidyverse)
library(sjPlot)
library(ggpubr)

# fix the code below to read in your cleaned data. delete this text afterward.
# load(here::here("..... ")) 
```

# Part 1

## Name of categorical variable 1 here
> Describe this variable in words (what does it say in the codebook) then delete this text. i.e., tell us what you're going to look at

```{r}
# put your code to create the summary table here
```

```{r}
# put your code to create the bar chart here
```

> Describe the distribution of the variable here in paragraph form. Then delete this text. 


## Name of categorical variable 2 here
> Describe this variable in words (what does it say in the codebook) then delete this text. i.e., tell us what you're going to look at

```{r}
# put your code to create the summary table here
```

```{r}
# put your code to create the bar chart here
```

> Describe the distribution of the variable here in paragraph form. Then delete this text. 


## Name of quantitative variable 1 here

> Describe this variable in words (what does it say in the codebook) then delete this text. i.e., tell us what you're going to look at

```{r}
# put your code to create the summary statistics
```

```{r}
#| layout-ncol: 2

# put your code to create the histogram AND boxplot here
```

> Describe the distribution of the variable here in paragraph form. Then delete this text. 




## Name of quantitative variable 2 here
> Describe this variable in words (what does it say in the codebook) then delete this text. i.e., tell us what you're going to look at


```{r}
# put your code to create the summary statistics
```

```{r}
#| layout-ncol: 2

# put your code to create the histogram AND boxplot here
```

> Describe the distribution of the variable here in paragraph form. Then delete this text. 

## Part II

> Use tbl summary to create a summary table for all 4 variables used in this assignment
