---
title: "Describing Relationships between variables"
date: "2022-09-19"
description: "5, 6"
author: "Robin Donatello"
footer: "[🔗 https://norcalbiostat.github.io/MATH615](https://norcalbiostat.github.io/MATH615/) / Describing  Relationships"
from: markdown+emoji
format: 
  revealjs:
    theme: beige
    multiplex: true
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
execute:
  freeze: auto
  echo: false
knitr:
  opts_chunk: 
    R.options:
      width: 200
---

# Bivariate: Association between 2 variables

## Naming conventions

|      Response      |     Explanatory      |
|:------------------:|:--------------------:|
|         y          |          x           |
|      outcome       |      predictor       |
| dependent variable | independent variable |
|                    |      covariate       |
|                    |       feature        |

\

Model notation: $y \sim x$

## Types of combinations

-   Categorical response and categorical explanatory variable. (C \~ C)
-   Quantitative response and categorical explanatory variable. (Q \~ C)
-   Quantitative response and quantitative explanatory variable. (Q \~ Q)

#  {background-image="images/penguins.png" background-size="60%"}

```{r}
#| echo: false
library(palmerpenguins); library(ggplot2); library(sjPlot); library(ggpubr)
pen <- penguins
```

## Categorical response & Categorical explanatory

::: panel-tabset
### Two way tables (cross-tabs)

::: columns
::: {.column width="50%"}
```{r}
#| echo: true
table(pen$species, pen$island)
```
:::

::: {.column width="50%"}
```{r}
#| echo: true
table(pen$species, pen$island) |> prop.table(margin=2) |> round(2)
```
:::
:::

-   All of the 52 penguins on Torgersen island are the *Adelie* species.
-   74% of penguins on Biscoe island are *Gentoo*.

### Plots

```{r}
plot_xtab(pen$island, grp=pen$species, show.total = FALSE)
```
:::

## Watch your margins {.smaller}

::: panel-tabset
### Tables

::: columns
::: {.column width="50%"}
**Row Percents**

```{r}
table(pen$species, pen$island) |> 
  prop.table(margin=1) |> round(2)
```

29% **of *Adelie* penguins** are on Biscoe Island.
:::

::: {.column width="50%"}
**Column Percents**

```{r}
table(pen$species, pen$island) |> 
  prop.table(margin=2) |> round(2)
```

74% **of penguins on Biscoe island** are *Gentoo*.
:::
:::

### Plots

::: columns
::: {.column width="50%"}
Distribution of islands for each species inhabits.

```{r}
plot_xtab(pen$species, grp=pen$island, margin = "row", show.total = FALSE)
```

29% **of *Adelie* penguins** are on Biscoe Island.
:::

::: {.column width="50%"}
Distribution of Species on each island.

```{r}
plot_xtab(pen$island, grp=pen$species, margin = "row", show.total = FALSE)
```

74% **of penguins on Biscoe island** are *Gentoo*.
:::
:::
:::

## Quantitative response & Categorical explanatory


::: panel-tabset
### Grouped summary statistics

### Plots

::: columns
::: {.column width="50%"}
```{r}
#| echo: true

```
:::

::: {.column width="50%"}
```{r}
#| echo: true

```
:::

:::
:::


## Quantitative response & Quantitative explanatory

::: panel-tabset
### Summary statistic

```{r}
#| echo: true
cor(pen$flipper_length_mm, pen$body_mass_g, use="pairwise.complete.obs")
```

The penguin flipper length (mm) has a strong positive correlation with body mass (g), r=0.87

### Plots

::: columns
::: {.column width="50%"}
```{r}
#| echo: true
ggscatter(pen, x="flipper_length_mm", y = "body_mass_g")
```
:::

::: {.column width="50%"}
```{r}
#| echo: true
ggscatter(pen, x="flipper_length_mm", y = "body_mass_g", add = "loess", conf.int = TRUE)
```
:::

<small> The relationship between flipper length and body mass in penguins is relatively linear, but there may be possible clustering on a third variable. There appears to be two groups below and above a flipper length of about 205mm. </small>
:::
:::
