---
title: "Describing Relationships between variables"
date: "2024-09-23"
description: "lec05a"
author: "Robin Donatello"
footer: "[🔗 https://math615.netlify.app](https://math615.netlify.app) / Describing Relationships"
from: markdown+emoji
format: 
  revealjs:
    theme: sky
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    code-fold: true
    code-summary: "Show the code"
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:
      width: 200
---

# Bivariate Association between 2 variables

## Naming conventions

|      Response      |     Explanatory      |
|:------------------:|:--------------------:|
|         y          |          x           |
|      outcome       |      predictor       |
| dependent variable | independent variable |
|                    |      covariate       |
|                    |       feature        |

\

Model notation: $y \sim x$

## Types of combinations

-   Categorical response and categorical explanatory variable. (C \~ C)
-   Quantitative response and categorical explanatory variable. (Q \~ C)
-   Quantitative response and quantitative explanatory variable. (Q \~ Q)

#  {background-image="images/penguins.png" background-size="60%"}

```{r}
#| code-fold: false
library(tidyverse); library(ggpubr); library(gtsummary)
library(palmerpenguins);  library(sjPlot)
pen <- penguins
```

## C \~ C {.smaller}
Categorical Response vs Categorical Explanatory

::: panel-tabset
### Two way tables (cross-tabs)

::: columns
::: {.column width="50%"}
```{r}
#| echo: true
table(pen$species, pen$island)
```
:::

::: {.column width="50%"}
```{r}
#| echo: true
table(pen$species, pen$island) |> 
  prop.table(margin=2) |> round(2)
```
:::
:::

-   All of the 52 penguins on Torgersen island are the *Adelie* species.
-   74% of penguins on Biscoe island are *Gentoo*.


### Plots

```{r}
#| echo: true
plot_xtab(pen$island, grp=pen$species, 
          show.total = FALSE)
```
:::

## Watch your margins 

:::{.callout-warning}
### Always double check your work 
One of the most common places for a mistake when creating a plot or a table between two categorical variables is by not paying close attention to the choice of denominator. And then also confirming the interpretation matches the table, which matches the plot. 
:::

## Watch your margins {.smaller}

::: panel-tabset
### Tables

::: columns
::: {.column width="50%"}
**Row Percents**

```{r}
#| echo: true
table(pen$species, pen$island) |> 
  prop.table(margin=1) |> round(2)
```

29% **of *Adelie* penguins** are on Biscoe Island.
:::

::: {.column width="50%"}
**Column Percents**

```{r}
#| echo: true
table(pen$species, pen$island) |> 
  prop.table(margin=2) |> round(2)
```

74% **of penguins on Biscoe island** are *Gentoo*.
:::
:::


### gtsummary

::: columns
::: {.column width="50%"}
Distribution of Islands within each species
```{r}
pen %>% select(island, species) %>%
  tbl_summary(by = "species")
```
:::

::: {.column width="50%"}
Distribution of Species on each Island
```{r}
pen %>% select(island, species) %>%
  tbl_summary(by = "island")
```
:::
:::


### Plots

::: columns
::: {.column width="50%"}
Distribution of islands for each species

```{r}
#| echo: true
plot_xtab(pen$species, grp=pen$island, 
          margin = "row", show.total = FALSE)
```

29% **of *Adelie* penguins** are on Biscoe Island.
:::

::: {.column width="50%"}
Distribution of species on each island.

```{r}
#| echo: true
plot_xtab(pen$island, grp=pen$species, 
          margin = "row", show.total = FALSE)
```

74% **of penguins on Biscoe island** are *Gentoo*.
:::
:::
:::

## Q \~ C {.smaller}

Quantitative Response vs Categorical Explanatory

::: panel-tabset
### Grouped summary statistics

n, nean, median, sd, IQR of the quantitative variable for each level of the categorical level.

```{r}
#| echo: true
#| eval: false
pen %>% group_by(species) %>% 
  summarize(n=n(), 
            mean = mean(bill_depth_mm, na.rm = TRUE), 
            median = median(bill_depth_mm, na.rm = TRUE), 
            sd = sd(bill_depth_mm, na.rm = TRUE), 
            IQR = IQR(bill_depth_mm, na.rm = TRUE))
```

```{r}
pen %>% group_by(species) %>% 
  summarize(n=n(), 
            mean   = mean(bill_depth_mm, na.rm = TRUE), 
            median = median(bill_depth_mm, na.rm = TRUE), 
            sd     = sd(bill_depth_mm, na.rm = TRUE), 
            IQR    = IQR(bill_depth_mm, na.rm = TRUE))
```

Gentoo penguins have lower average bill depth compared to Adelie or Chinstrap (15.0mm vs 18.3 and 18.4mm respectively). Chinstrap however have a larger IQR at 1.9 compared to 1.5 for the others.

### Plots

::: columns
::: {.column width="50%"}
**Overlaid density plots**

```{r}
#| echo: true
gghistogram(pen, 
    x = "bill_depth_mm", fill = "species", 
    add_density = TRUE, add="mean")
```
:::

::: {.column width="50%"}
**Side by side boxplots**

```{r}
#| echo: true
ggviolin(pen, 
  x="species", y = "bill_depth_mm", 
  color = "species", add = c("mean", "boxplot"))
```
:::
:::

The distribution of bill depth are fairly normal for each species, with some higher end values causing a slight right skew for Adelie and Gentoo.

### Plot2
More code but nice alternative to violins. 
```{r}
library(ggdist) # for the "half-violin" plot (stat_slab)
ggplot(pen, aes(x=bill_depth_mm, y=species, fill=species)) + 
      stat_slab(alpha=.5, justification = 0) + 
      geom_boxplot(width = .2,  outlier.shape = NA) + 
      geom_jitter(alpha = 0.5, height = 0.05) +
      stat_summary(fun="mean", geom="point", col="red", size=4, pch=17) + 
      theme_bw() + 
      labs(x="Bill depth (mm)", y = "Species") + 
      theme(legend.position = "none")
```



:::

## Q \~ Q {.smaller}
Quantitative Response vs Quantitative Explanatory 

::: panel-tabset
### Summary statistic 

```{r}
#| echo: true
cor(pen$flipper_length_mm, pen$body_mass_g, 
    use="pairwise.complete.obs")
```

The correlation coefficient $r$ is a measure of the strength and direction of a linear relationship between two variables. 

-   $|r| > 0.6$ Strong relationship
-   $0.4 \leq |r| < 0.6$ Moderate relationship
-   $|r| < 0.4$ Weak to no relationship

Note: These values are guidelines and subject to variability within different disciplines [ref](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6107969/)

### Plots

::: columns
::: {.column width="50%"}
```{r}
#| echo: true
ggscatter(pen, 
  x="flipper_length_mm", y = "body_mass_g")
```
:::

::: {.column width="50%"}
```{r}
#| echo: true
ggscatter(pen, 
  x="flipper_length_mm", y = "body_mass_g", 
  add = "loess", conf.int = TRUE)
```
:::

<small> The relationship between flipper length and body mass in penguins is relatively linear, but there may be possible clustering on a third variable. There appears to be two groups below and above a flipper length of about 205mm. </small>
:::

### Description

Must include the `direction` (positive or negative), the `strength` quantified using $r$, and the form (linear vs non-linear)

> The penguin flipper length (mm) has a strong, positive, and reasonably linear correlation with body mass (g), r=0.87.

:::
