---
title: "Describing Distributions of Data"
date: "2022-09-12"
description: "4"
author: "Robin Donatello"
footer: "[🔗 https://norcalbiostat.github.io/MATH615](https://norcalbiostat.github.io/MATH615/) / Describing  Distributions of Data"
from: markdown+emoji
format: 
  revealjs:
    theme: beige
    multiplex: true
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    preview-links: true
execute:
  freeze: auto
  echo: false
knitr:
  opts_chunk: 
    R.options:
      width: 200
---

## Motivation
Visualizing your data is hands down the most important thing you can learn to do. 

* Screen for data entry errors
* Out of range variables
* Mistakes in coding
* Violation of model assumptions

## Level of care depends on the audience {.smaller}

There are three main audiences in mind when creating data visualizations: 

1. For your eyes only (FYEO). These are quick and dirty plots, without annotation. Meant to be looked at once or twice. 
    - You'll create a TON of these. Don't spend a ton of time on them. 
    
. . . 

2. To share with others internally. These mostly need to stand on their own. Axes labels, titles, colors as needed, possibly captions.
    - You'll create a lot of these, and you'll get better with practice at adding necessary annotation without a lot of time
    
. . . 

3. Professional - Contains all bells and whistles needed to make it publication quality. 
    - You'll create very few of these, but they demand a lot of time, detail and thought. 

## Graphing with intent

Along with having the audience in mind, it is important to give thought to the purpose of the chart.

> The effectiveness of any visualization can be measured according to how well it fulfills the tasks it was designed for. (A. Cairo, 2018).


## Choosing Appropriate Visualization

75% of your choice is determined by the data type

[https://r-graph-gallery.com/](https://r-graph-gallery.com/)

# Meet the Palmer Penguins {background-image="images/penguins.png" background-size="90%" background-position="right"}
```{r}
#| echo: false
library(palmerpenguins); library(ggplot2)
pen <- penguins
```

## Single Categorical {background-image="images/penguins.png" background-size="20%" background-position="80% 10%"}

::: {.panel-tabset}
### Tables

**Frequencies (N)**
```{r}
#| echo: true
table(pen$species)
```

**Percents (%)**
```{r}
#| echo: true
table(pen$species) |> proportions() |> round(digits=2)
```

Penguin species _Adelie_ make up 44% of the sample (n=152)

### Plots

```{r}
#| echo: true
#| fig.height:  4
sjPlot::plot_frq(pen$species) + xlab("Species")
```
:::

## Single Numeric {background-image="images/culmen_depth.png" background-size="25%" background-position="100% 0%"}


::: {.panel-tabset}
### Center
```{r}
#| echo: true
summary(pen$bill_depth_mm)
sd(pen$bill_depth_mm, na.rm=TRUE)
```

\

The average bill depth is 17.15mm, with a standard deviation of 1.9mm. 

### Spread

```{r}
#| echo: true
summary(pen$bill_depth_mm)
IQR(pen$bill_depth_mm, na.rm=TRUE)
```
\

Bill depth ranges from 13.1 to 21.5mm, and has an IQR of 3.1mm.


### Plots

:::: {.columns}

::: {.column width="50%"}
```{r}
#| echo: true
ggpubr::gghistogram(pen$bill_depth_mm, add_density = TRUE)
```

:::

::: {.column width="50%"}
```{r}
#| echo: true
ggpubr::ggviolin(pen$bill_depth_mm, add = c("jitter", "boxplot")) + coord_flip()
```

:::

::::

The distribution of bill depth appears to be bimodal with peaks around 15 and 18mm. 

:::

## How to create graphs

* Similar to the data management section, after identifying what you want to do, you go look up how to do that thing. 
* Don't expect to remember the exact code yet, just know where to look up an example and copy from there
    - Prior semester SPSS students crated a HackMD collaborative notes file for sharing code. [[LINK]](https://hackmd.io/@norcalbiostat/SPSS)
    - Need these for other languages!
* copy/paste/pray
* Keep graphs simple until you get more comfortable. 


## References

* PMA6 Chapter 4 for appropriate plot choices
* Applied Stats course Notes Chapter 2.3 for examples and code
* HackMD Notes
