---
title: "Inference on proportions"
date: "2025-10-20"
author: "Robin Donatello"
description: "lec07d"
from: markdown+emoji
format: 
  revealjs:
    theme: sky
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:
      width: 200
---

```{r, warning=FALSE, message=FALSE, include=FALSE}
library(knitr)
library(tidyverse)
library(scales)
library(sjPlot)
library(ggpubr)
library(gtsummary)

source(here::here("load_data.R"))
```

## Opening Remarks

If we are only concerned with testing the hypothesis that the proportion
of successes between two groups are equal $p_{1}-p_{2}=0$, we can
leverage the Normal distribution and conduct a Z-test for two
proportions.

However in this class we will use the more generalizable model via
Chi-squared test of association/equal proportions.

## Chi-squared test of equal proportions

A 30-year study was conducted with nearly 90,000 female participants.
(Miller AB. 2014) During a 5-year screening period, each woman was
randomized to one of two groups: in the first group, women received
regular mammograms to screen for breast cancer, and in the second group,
women received regular non-mammogram breast cancer exams. No
intervention was made during the following 25 years of the study, and
we'll consider death resulting from breast cancer over the full 30-year
period.

## Results from study

```{r, echo=FALSE}
a <- data.frame(Tx = c(rep("Mammogram", 500), rep("Control", 505), 
                  rep("Mammogram", 44425), rep("Control", 44405)),
           Outcome = c(rep("Dead",1005), rep("Alive", 88830)))

addmargins(table(a$Tx, a$Outcome))%>% kable() 
```

The explanatory variable is treatment (additional mammograms), and the response variable is death from breast cancer. 

Are these measures associated?

## Assume independence/no association {.smaller .incremental}

-   If mammograms are more effective than non-mammogram breast cancer
    exams, then we would expect to see additional deaths from breast
    cancer in the control group (there is a relationship).
-   If mammograms are not as effective as regular breast cancer exams,
    we would expect to see no difference in breast cancer deaths in the
    two groups (there is no relationship).
-   Need to figure out how many deaths would be **expected**, if there
    was no relationship between treatment death by breast cancer,
-   Then examine the **residuals** - the difference between the observed
    counts and the expected counts in each cell.

## Table notation

Tables can be described by $i$ rows and $j$ columns. So the cell in the
top left is $i=1$ and $j=1$.

| $O_{ij}$ |  Alive   |         Dead         |        Total         |
|:---------|:--------:|:--------------------:|:--------------------:|
| Mammo    | $n_{11}$ |       $n_{12}$       |      $n_{1.}$        |
| Control  | $n_{21}$ |       $n_{22}$       |      $n_{2.}$        |
| Total    | $n_{.1}$ |       $n_{.2}$       |         $N$          |

In our DATA = MODEL + RESIDUAL framework, the DATA ($n_{11}, n_{12}$,
etc)is the observed counts $O_{ij}$ and the MODEL is the expected counts
$E_{ij}$.

## Calculating the expected count

Since we assume the variables are independent (unless the data show
otherwise) the expected count for each cell is calculated as the row
total times the column total for that cell, divided by the overall
total.

$$E_{ij} = \frac{n_{i.}n_{.j}}{N}$$

In a probability framework this is stated as two variables $A$ and $B$
are independent if
$P(A \cap B) = P(A)*P(B) = \frac{n_{i.}}{N}*\frac{n_{.j}}{N}$

## Calculating the residuals {.smaller}

The residuals are calculated as $(O_{ij} - E_{ij})$, observed minus expected. 

::: columns
::: {.column width="50%"}
Expected

```{r}
chisq.test(a$Tx, a$Outcome)$expected %>% 
  kable(digits=2)
```
:::

::: {.column width="50%"}
Residuals

```{r}
chisq.test(a$Tx, a$Outcome)$residuals %>% 
  kable(digits=3) 
```
:::
:::

If mammograms were not associated with survival, there were 0.01 fewer
people still alive than expected, and 0.11 more people dead.

## $\chi^2$ test statistic {.smaller}

::: columns
::: {.column width="50%"}
The $\chi^2$ test statistic is defined as the sum of the squared
residuals, divided by the expected counts, and follows a $\chi^2$
distribution with degrees of freedom (#rows -1)(#cols -1).

$$ \sum_{ij}\frac{(O_{ij}-E_{ij})^{2}}{E_{ij}} $$

**Assumptions for this distribution to hold**

* Sample size is large
* $E_{ij} \geq 5$ for at least 80% of cells
:::

::: {.column width="50%"}
```{r}
chisq.test(a$Tx, a$Outcome)
dist_chisq(chi2=.01748, deg.f=1, xmax = 10,
           geom.colors=c("#FDE725FF", "#440154FF"))
```
:::
:::


## Conclusion

In this example, the test statistic was
`r round(chisq.test(a$Tx, a$Outcome)$statistic,3)` with p-value
`r round(chisq.test(a$Tx, a$Outcome)$p.value,3)`.

So there is not enough reason to believe that mammograms in addition to
regular exams are associated with a reduced risk of death due to breast
cancer.

This example demonstrated how we examine the residuals to see how
closely our DATA fits a hypothesized MODEL.

## Example: Income and General Health

Using the Addhealth data set, what can we say about the relationship
between smoking status and a person's perceived level of general health?

Is there an association between lifetime smoking status and perceived
general health?

## 1. Identify response and explanatory variables

-   The binary explanatory variable is whether the person has ever
    smoked an entire cigarette (`eversmoke_c`).
-   The categorical explanatory variable is the person's general health
    (`genhealth`) and has levels "Excellent", "Very Good", "Good",
    "Fair", and "Poor".

## 2. Visualize and summarise {.smaller}

::: {.panel-tabset}
## Standard


::: columns
::: {.column width="55%"}
```{r}
plot_xtab(grp=addhealth$eversmoke_c, x=addhealth$genhealth, 
                  show.total = FALSE, margin="row", legend.title="") + 
   theme_bw() + theme(legend.position="bottom") + xlab("General Health")
```
:::

::: {.column width="45%"}
The percentage of smokers seems to increase as the general health status
decreases. Almost three-quarters (73%, n=40) of those reporting poor
health have smoked an entire cigarette at least once in their life
compared to 59% (n=573) of those reporting excellent health.
:::
:::


## Alternative visualizations

::: columns
::: {.column width="50%"}
Plot the % who smoked at least once. This method only works when there is a binary outcome. 
```{r}
#| fig-height: 4
addhealth %>% 
  group_by(genhealth) %>% 
  summarize(p = mean(eversmoke_c == "Smoked at least once", na.rm=TRUE)*100, 
            n = n()) %>% 
  na.omit() %>% 
  ggplot(aes(x=genhealth, y=p, color=genhealth)) + 
  geom_point(aes(size = n)) + 
  scale_y_continuous(limits=c(0, 100)) + 
  geom_segment(aes(x=genhealth, xend=genhealth, y=0, yend=p)) +
  scale_color_discrete(guide="none") + theme_bw() + 
  ylab("Proportion of smokers") + 
  xlab("Perceived General Health")
```
:::

::: {.column width="50%"}
A stacked barchart can work for any size categorical outcome, but the more categories/segments, the harder it will be to see the %'s. 
```{r}
#| fig-height: 4
addhealth %>% 
  count(genhealth, eversmoke_c) %>% 
  na.omit() %>%
  group_by(genhealth)%>%
  mutate(p = n/sum(n)*100) %>% 
ggplot(aes(x=genhealth, y=p, fill=eversmoke_c, 
           label = paste0(round(p, 1), "%"))) + 
  geom_col() + theme_bw() + 
  scale_fill_discrete(name = "") + 
  geom_text(position=position_stack(0.5)) + 
  ylab("Proportion of smokers") + 
  xlab("Perceived General Health")
```
:::
:::

:::


## 3.  Write the null and research hypothesis in words and symbols. {.smaller}

-   Null Hypothesis: The proportion of smokers is the same across all
    levels of general health.
-   Alternate Hypothesis: At least one group has a different proportion
    of smokers compared to the other general health groups.

Let $p_{1}, p_{2}, \ldots, p_{5}$ be the true proportion of smokers in
each of the 5 health categories: (1) Excellent to (5) Poor.

\

$H_{0}: p_{1} = p_{2} = \ldots p_{5}$\
$H_{A}:$ At least one $p_{j}$ is different.

## 4. State and justify the analysis model. Check assumptions. {.smaller}

* I will conduct a $\chi^{2}$ test of association because both variables are categorical. 
* The expected value in each combination of smoking status and general health is at least 5. 

```{r}
chisq.test(addhealth$genhealth, addhealth$eversmoke_c)$expected
```

## 5. Conduct the test {.smaller}
And make a decision about the plausibility of the alternative hypothesis.

```{r}
chisq.test(addhealth$genhealth, addhealth$eversmoke_c)
```

We have strong evidence in favor of the alternative hypothesis, p\<.0001

### 6. Write a conclusion

We can conclude that there is an association between ever smoking a cigarette in their life and perceived general health ($\chi^{2} = 30.8, df=4, p<.0001$).

## Multiple comparisons: Which group is different? 

Just like with ANOVA, if we find that the Chi-squared test indicates
that at least one proportion is different from the others, it's our job
to figure out which ones might be different! We will analyze the
residuals to accomplish this.

## Examine the residuals {.smaller}

The residuals ($O_{ij} - E_{ij}$) are automatically stored in the model
output. You can either print them out and look at the values directly or
create a plot. You're looking for combinations that have much higher, or
lower expected proportions.

::: columns
::: {.column width="50%"}
```{r}
health.smoke.model <- chisq.test(addhealth$genhealth, addhealth$eversmoke_c)
health.smoke.model$residuals |> round(2)
```
:::

::: {.column width="50%"}
```{r}
#| fig-height: 4
plot.residuals <- health.smoke.model$residuals %>% data.frame()
ggplot(plot.residuals, aes(x=addhealth.genhealth, y=addhealth.eversmoke_c)) +
       geom_raster(aes(fill=Freq)) +  scale_fill_viridis_c()
```
:::
:::

The proportion of non-smokers in the excellent health category is much
higher than expected if there were no relationship between these two
variables.

## All pairwise comparisons {.smaller}

::: columns
::: {.column width="70%"}
```{r}
# library(rcompanion)
smoke.by.genhealth <- table(addhealth$genhealth, addhealth$eversmoke_c)
rcompanion::pairwiseNominalIndependence(smoke.by.genhealth, fisher = FALSE, gtest=FALSE) |> kable(digits = 3)
```
:::

::: {.column width="30%"}
Using the adjusted (for multiple comparisons) p-value column, the
proportion of smokers in the Excellent group (41.3%) is significantly
different from nearly all other groups (35.4% for Very Good, 31.3% for
Good, 31.8% for Fair).

But why is it not significantly different from the 27.3% in the Poor
group?
:::
:::

## The effect of small sample sizes {.smaller}

Why is 41.3% significantly different from 35.4%, but NOT different from 27.3%?

The standard error is always dependent on the sample size. Here, the number of individuals in the Poor health category is much smaller compared to all other groups - so the margin of error will be larger - making it less likely to be statistically significantly different from other groups.

When you have a small numbers (typically n\<10) in one or more cells the value of the expected value $E_{ij}$ for that cell likely will be below, or close to 5. This means distribution of the test statistic can no longer be modeled well using the $\chi^{2}$ distribution. So we again go to a non-parametric test that uses a randomization based method.

## Fishers Exact Test {.smaller}

```{r}
fisher.test(addhealth$genhealth, addhealth$eversmoke_c, 
            simulate.p.value = TRUE)
```

Then you can use the `pairwiseNominalIndependence` function for all
pairwise comparisons again, with the `fisher` argument set to `TRUE`.

::: {.callout-note appearance="minimal"}
<small>Note: the `simulate.p.value = TRUE` argument here is only needed
if the randomization space is larger than your workspace. R will let you
know when this is needed.</small>
:::

# Chi-Squared test of Association or Independence

## C ~ C

We can still use the $\chi^{2}$ test of association to compare categorical variables with more than 2 levels. In this case we generalize the statement to ask: **Is the distribution of 1 variable the same across levels of another variable?**. In this sense, it is very much like an ANOVA.

Mathematically the $\chi^{2}$ test of association is the exact same as a test of equal proportions.

## Example

Let's analyze the relationship between a person's income and perceived level of general health.

The categorical explanatory variable is income, binned into 4 ranges
(`income_cat`).

```{r}
addhealth$income_cat <- Hmisc::cut2(addhealth$income, g = 4)
```

## Visualize and summarize {.smaller}

```{r}
#| fig.height: 4
plot_xtab(grp=addhealth$income_cat,
                  x=addhealth$genhealth, 
                  show.total = FALSE, margin="row",
                  legend.title="") + 
  theme(legend.position = "top")
```

The income distribution is nearly flat for who rate themselves
in excellent or very good condition. However, the proportion of
individuals in the lower income categories increase as perceived general
health decreases. Over 80% of those that rate themselves as in Poor
condition have an annual income less than \$35,500.

## State Hypothesis {.smaller}

Let $p_{1e}, p_{2e}, p_{3e}, p_{4e}$ be the true proportion of income brackets within those that say they are in _excellent_ health. 

Let $p_{1v}, p_{2v}, p_{3v}, p_{4v}$ be the true proportion of income brackets within those that say they are in _very good_ health. 

... and so forth for each health category. 

$H_{0}:$ The income distribution ($p_{1j}, p_{2j}, p_{3j}, p_{4j}$) is the same in each of the $j$ general health categories (no association)

$H_{A}:$ The income distribution differs for at least one of the general health categories (association)


## Test the relationship & make a conclusion. {.smaller}

```{r}
chisq.test(addhealth$genhealth, addhealth$income_cat)
chisq.test(addhealth$genhealth, addhealth$income_cat)$expected
```

There is sufficient evidence to conclude that the distribution of income
is not the same across all general health status categories. The Chi-squared test of association is valid here because of the large sample size, and all expected values are over 5. 

## Other analysis for categorical variables

There is a slew of methods to analyze categorical data. To learn more start with Categorical Data Analysis by Alan Agresti, who has written *extensively* on the subject.



