---
title: "Logistic Regression"
date: "2025-10-10"
author: "Robin Donatello"
description: "lec10b"
footer: "[🔗 https://math615.netlify.app](https://math615.netlify.app)"
from: markdown+emoji
format: 
  revealjs:
    theme: sky
    transition: fade
    slide-number: true
    incremental: false 
execute:
  freeze: auto
  echo: true
  message: false
  warning: false
knitr:
  opts_chunk: 
    R.options:
      width: 200
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| echo: false
library(gtsummary)
library(survey)
library(ggpubr)
library(ggplot2)
load(here::here("data/addhealth_clean.Rdata"))
depress <- read.delim(here::here("data/Depress.txt"))
names(depress) <- tolower(names(depress))
```

## Binary outcome data

Consider an outcome variable $Y$ with two levels: Y = 1 if event, = 0 if no event. 

Let $p_{i} = P(y_{i}=1)$. 

Two goals: 

* Assess the impact selected covariates have on the probability of an outcome occurring. 
* Predict the probability of an event occurring given a certain covariate pattern. 

Binary data can be modeled using a _Logistic Regression Model_ 

## What are the Odds? 

The _odds_ are defined as the probability an event occurs divided by the probability it does not occur: $\frac{p_{i}}{1-p_{i}}$. 

\

The function $ln\left(\frac{p_{i}}{1-p_{i}}\right)$ is also known as the _log odds_, or more commonly called the **_logit_**. This is the _link_ function for the logistic regression model. 

## Link function
We use this `logit` function to transform a binary outcome (only 0 or 1) variable into a continuous probability (which only has a range from 0 to 1). 
```{r}
p <- seq(0, 1, by=.01)
logit.p <- log(p/(1-p))
qplot(logit.p, p, geom="line", xlab = "logit(p)", main="The logit transformation") + 
  theme_bw()
```


## Logistic Regression

The logistic model then relates the probability of an event based on a linear combination of X's. 

$$
log\left(
\frac{p_{i}}{1-p_{i}}
\right) = \beta_{0} + \beta_{1}x_{1i} + \beta_{2}x_{2i} + \ldots + \beta_{p}x_{pi}
$$

This means the relationship between $X$ and the probability of success is nonlinear, but the relationship between $X$ and the log-odds is linear. 

## Deriving the Odds Ratio (OR)

* Logistic regression model: $logit(y) = \beta_0 + \beta_1 X$

* The odds at $X = x$ is $e^{\beta_0 + \beta_1 x}$

* The odds at $X = x+1$ is $e^{\beta_0 + \beta_1 (x+1)} = e^{\beta_0 + \beta_1 x} * e^{\beta_1}$

* The **odds ratio (OR) for a 1 unit change in $X$ is then $e^{\beta_1}$

The OR measures how the **odds of success** change for a one-unit increase in $X$, holding other variables constant.

## Interpreting the OR

Consider a binary outcome with values YES, coded as 1, and NO, coded as 0. 

* **OR = 1** = equal chance of response variable being YES given any explanatory variable value. 
* **OR > 1** = as the explanatory variable value increases, the presence of a YES response is more likely. 
* **OR <1** = as the explanatory variable value increases, the presence of a YES response is less likely.

## Confidence Intervals {.smaller}

* The OR is **not** a linear function of the $x's$, but $\beta$ is. 
* This means that a CI for the OR is created by calculating a CI for $\beta$, and then exponentiating the endpoints.
* A 95% CI for the OR is calculated as: 

$$e^{\hat{\beta} \pm 1.96 SE_{\beta}} $$

This math holds for any $k$ unit change in x. The linearity of the confidence interval only applies at the untransformed level of the $\beta$'s. NOT the odds ratio. 


## Example: Depression {.smaller}

Let's fit a model to examine the effect of identifying as female (`gender`) has on a depression (`cases`) diagnosis. 

```{r}
dep_sex_model <- glm(cases ~ sex, data=depress, family="binomial")
summary(dep_sex_model)
```

## Calculate the Odds Ratio

We exponentiate the coefficients to back transform the $\beta$ estimates into Odds Ratios
```{r}
tbl_regression(dep_sex_model, exponentiate = TRUE)
```

Females have 2.8 (1.4, 6.2) times the odds of showing signs of depression compared to males (p = 0.006). 

::: {.callout-important}
note the multiplicative effect language "times the odds", not just "higher odds"
:::
 

## Multiple Logistic Regression {.smaller}
Let's continue with the depression model, but now also include age and income as potential predictors of symptoms of depression. 

```{r}
mvmodel <- glm(cases ~ age + income + sex, data=depress, family="binomial")
tbl_regression(mvmodel, exponentiate = TRUE)
```

* The odds of a female being diagnosed with depression is 2.53 (1.23, 5.66) times greater than the odds for Males after adjusting for the effects of age and income (p=.016). 


## Model Fit 

* **Pseudo $R^{2}$** Not appropriate for logistic regression
* **Hosmer and Lemeshow (1980) "Goodness of Fit"** take a "measure the residuals" approach to estimate how well the model fits the data. 
    * Implemented in the R package: `MKmisc`, function `HLgof.test`
* **Prediction Accuracy** Using the model to calculate predicted probabilties of $Y=1$, how often does the model prediction match the data? 

