---
title: "Generalized Linear Models"
date: "2025-11-10"
author: "Robin Donatello"
description: "lec10a"
footer: "[🔗 https://math615.netlify.app](https://math615.netlify.app)"
from: markdown+emoji
format: 
  revealjs:
    theme: sky
    transition: fade
    slide-number: true
    incremental: false 
execute:
  freeze: auto
  echo: true
  message: false
  warning: false
knitr:
  opts_chunk: 
    R.options:
      width: 200
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| echo: false
library(gtsummary)
library(ggpubr)
source(here::here("load_data.R"))
```

## Introduction
One of the primary assumptions with linear regression, is that the error terms have a specific distribution. Namely: 

$$ \epsilon_{i} \sim \mathcal{N}(0, \sigma^{2}) \qquad i=1, \ldots, n, \quad \mbox{and } \epsilon_{i} \perp \epsilon_{j}, i \neq j $$

When your outcome variable $y$ is non-continuous/non-normal, the above assumption fails dramatically. 

## Generalized Linear Models (GLM) {.smaller}

Allows for different data type outcomes by allowing the linear portion of the model ($\mathbf{X}\beta$) to be related to the outcome variable $y$ using a _link_ function, that allows the magnitude of the variance of the errors ($\sigma$) to be related to the predicted values themselves. 

There are a few overarching types of non-continuous outcomes that can be modeled with GLM's. 

* Binary data: Logistic or Probit regression
* Log-linear models 
* Multinomial/categorical data: Multinomial or Ordinal Logistic regression.  
* Count data: Poisson regression 


## Linking the response to the predictors {.smaller}

All regression models aim to model the expected value of the response variable $Y$ given the observed data $X$, through some link function $C$ 

$$E(Y|X) = C(X)$$ 

Depending on the data type of $Y$, this link function takes different forms. Examples include: 

* Linear regression: C = Identity function (no change)
* Logistic regression: C = logit function
* Poisson regression: C = log function

## Fitting GLM

The general syntax is similar to `lm()`, with the additional required `family=` argument. See `?family` for a list of options. 

Example for Logistic regression would be: 

```{r}
#| eval: false
#| code-fold: false
glm(y ~ x1 + x2 + x3, data=DATA, family="binomial") 
```

## Log-linear models {.smaller}

A *log-linear* model is when the log of the response variable is modeled using a linear combination of predictors. 

$$ln(Y) \sim XB +\epsilon$$

* In statistics, when we refer to the _log_, we mean the natural log _ln_.
* This type of model is often use to model count data using the Poisson distribution, or to achieve normality when the response variable is right skewed. 

## Interpreting results

Since we transformed our outcome before performing the regression, we have to back-transform the coefficient before interpretation. Similar to logistic regression, we need to _exponentiate_ the regression coefficient before interpreting. 

When using log transformed outcomes, the effect on Y becomes **multiplicative** instead of additive. 

* **Additive** For every 1 unit increase in X, y increases by b1
* **Multiplicative** For every 1 unit increase in X, y is multiplied by $e^{b1}$

## Example

let $b_{1} = 0.2$. 

* **Additive** For every 1 unit increase in X, y increases by 0.2 units.
* **Multiplicative** For every 1 unit increase in X, y changes by a factor of $e^{0.2} = 1.22$ 

Sign of coefficient still is directly connected to direction of association. 

## Percent Change {#sec-pct-chg .smaller}

Thus we interpret the coefficient as a **percentage** change in $Y$ for a unit increase in $x_{j}$. We are interested in the difference between the estimate and the value of 1. 

* Let $b_{j} = -0.2$. Then $e^{-0.2} = 0.82$, and $(1-0.82) = .18$
    * The expected value of $Y$ for when $x=1$ is 18% _lower_ than when $x=0$
* Let $b_{j} = 0.2$. Then $e^{0.2} = 1.22$, and $(1.22-1) = .22$
    * The expected value of $Y$ for when $x=1$ is 22% percent _higher_ than when $x=0$
   

## Example: Personal Income {.smaller} 

We are going to analyze personal income from the AddHealth data set. Income naturally is right skewed, but a log transformation fixes this problem nicely. 
```{r}
#| layout-ncol: 2
#| layout-nrow: 2
#| fig-height: 2
#| fig-width: 3
ggdensity(addhealth, x = "income", fill = "springgreen4") +
   stat_overlay_normal_density(color = "darkgreen", linetype = "dashed")
ggdensity(addhealth, x = "logincome", fill = "skyblue") +
   stat_overlay_normal_density(color = "navy", linetype = "dashed")

ggqqplot(addhealth, x = "income")
ggqqplot(addhealth, x = "logincome")

```

## Identify variables

* Quantitative outcome that has been log transformed: Income (variable `logincome`)
* Quantitative predictor: typical time waking up on a work day  (variable `wakeup`)
* Binary predictor: Gender (variable `female_c`)
 
The mathematical multivariable model looks like: 

$$ln(Y) \sim \beta_{0} + \beta_{1}x_{1} + \beta_{2}x_{2}$$

## Fit a linear regression model {.smaller}

```{r}
#| code-fold: false
ln.mod.2 <- lm(logincome~wakeup + female_c, data=addhealth)
summary(ln.mod.2)
```

The fitted model is: $ln(\hat{y}) = 10.65 - 0.0149x_{wakeup} -0.193x_{female}$

The coefficient estimates $b_{p}$ are not interpreted directly as an effect on $y$. 

## Exponentiate & Interpret

Recap: $e^{b_p}$ is the percent change. (ref @sec-pct-chg)

* For every hour later one wakes up in the morning, one can expect to earn `1-exp(-0.015)` = 1.4% less income than someone who wakes up one hour earlier. This is after controlling for gender. 
* Females have on average `1-exp(-0.19)` = 17% percent lower income than males, after controlling for the wake up time. 

## All the math at once {.smaller}

Extract point estimate using `coef()`
```{r}
#| code-fold: false
exp(coef(ln.mod.2))
```

Extract CI using `confint()`
```{r}
#| code-fold: false
1-exp(confint(ln.mod.2)[-1,])
```

Both gender and time one wakes up are significantly associated with the amount of personal earnings one makes. Waking up later in the morning is associated with 1.4% (95% CI 0.8%-2%, p<.0001) percent lower income than someone who wakes up one hour earlier. Females have 17% (95% CI 15%-20%, p<.0001) percent lower income than males. 