---
title: "Classification"
date: "2025-11-12"
author: "Robin Donatello / MATH 615"
description: "lec10c"
footer: "[🔗 https://math615.netlify.app](https://math615.netlify.app)"
from: markdown+emoji
format: 
  revealjs:
    theme: sky
    transition: fade
    slide-number: true
    incremental: false 
execute:
  freeze: auto
  echo: true
  message: false
  warning: false
knitr:
  opts_chunk: 
    R.options:
      width: 200
editor: 
  markdown: 
    wrap: 72
---

## Setup

```{r}
#| code-fold: show
library(ggplot2)
library(ROCR)
library(ggpubr)
library(caret)
library(gtsummary)
source(here::here("load_data.R"))
```

## Predicted Probabilities {.smaller}

-   Sometimes Odds Ratios can be difficult to interpret or understand.
-   Sometimes you just want to report the probability of the event
    occurring.
-   Or sometimes you want to predict whether or not a new individual is
    going to have the event.

For all of these, we need to calculate $p_{i} = P(y_{i}=1)$, the
probability of the event.

$$
p_{i} = \frac{e^{\beta_{0} + \beta_{1}x_{1i} + \beta_{2}x_{2i} + \ldots + \beta_{p}x_{pi}}}
{1 + e^{\beta_{0} + \beta_{1}x_{1i} + \beta_{2}x_{2i} + \ldots + \beta_{p}x_{pi}}}
$$

## Example: {.smaller}

```{r, echo=FALSE}
depress$sex <- factor(depress$sex, labels = c("Male", "Female"))
dep_sex_model <- glm(cases ~ age + income + sex, data=depress, family="binomial")
tbl_regression(dep_sex_model)
```

The predicted probability of depression is calculated as: $$
P(depressed) = \frac{e^{-0.676 - 0.02096*age - .03656*income + 0.92945*sex}}
{1 + e^{-0.676 - 0.02096*age - .03656*income + 0.92945*sex}}
$$

Notice this formulation requires you to specify a *covariate profile*. In other words, what value X take on for each record. Often when you are only concerned with comparing the effect of a single measures, you set all other measures equal to their means.

## Comparing probabilities {.smaller}

Let's compare the probability of being depressed for males and females separately, while holding age and income constant at the average value calculated across all individuals (regardless of sex).

```{r}
#| eval: false
#| echo: false
XB.f <- -0.676 - 0.02096*(44.4) - .03656*(20.6) + 0.92945
phat.f <- exp(XB.f) / (1+exp(XB.f))
XB.m <- -0.676 - 0.02096*(44.4) - .03656*(20.6)
phat.m <- exp(XB.m) / (1+exp(XB.m))
```

$$
P(depressed|Female) = \frac{e^{-0.676 - 0.02096(44.4) - .03656(20.6) + 0.92945(1)}}
{1 + e^{-0.676 - 0.02096(44.4) - .03656(20.6) + 0.92945(1)}} = 0.193
$$


$$
P(depressed|Male) = \frac{e^{-0.676 - 0.02096(44.4) - .03656(20.6) + 0.92945(0)}}
{1 + e^{-0.676 - 0.02096(44.4) - .03656(20.6) + 0.92945(0)}} = 0.086
$$


The probability for a 44.4 year old female who makes \$20.6k annual income has a 0.19 probability of being depressed. The probability of depression for a male of equal age and income is 0.086.

## Distribution of Predicted probabilities {.smaller}

We can use the `predict()` command to generate a vector of predictions $\hat{p}_{i}$ for each row used in the model.

```{r}
phat.depr <- predict(dep_sex_model, type='response') 
gghistogram(phat.depr, add = "mean" , fill = "skyblue")
```

The average predicted probability of showing symptoms of depression is 0.17.

## Plotting predictions {.smaller}

There was a significant effect of gender on depression diagnosis. How does that appear in the predicted probabilities? 

::: {.callout-caution title="Handling missing data in the model"}
Any row with missing data on any variable used in the model will be dropped, and so NOT get a predicted value. So the tactic is to use the data stored in the model object.
:::

```{r}
#| code-fold: show
model.pred.data <- cbind(dep_sex_model$data, phat.depr)
tail(names(model.pred.data))
```

##  Model predicted probability of depression

:::: {.columns}

::: {.column width="60%"}
```{r}
ggpubr::ggdensity(model.pred.data, x="phat.depr", add="mean", color = "sex", fill = "sex", palette = c("#00AFBB", "#E7B800"))
```
:::

::: {.column width="40%"}
-   Reminder this is the fitted data $\hat{p}$ 
*   What do you notice in this plot?
:::

::::


## Predicted Class (outcome)

To classify individual $i$ as being depressed or not, we draw a binary value ($x_{i} = 0$ or $1$), with probability $p_{i}$ by using the `rbinom` function, with a `size=1`.

```{r}
set.seed(12345) #reminder: change the combo on my luggage
plot.mpp <- data.frame(pred.prob = phat.depr, 
                       pred.class = rbinom(n = length(phat.depr), 
                                           size = 1, 
                                           p = phat.depr),
                       truth = dep_sex_model$y)
head(plot.mpp)
```

## Confusion Matrix

Applying class labels and creating a cross table of predicted (rows) vs truth (cols):

```{r}
plot.mpp <- plot.mpp %>% 
            mutate(pred.class = factor(pred.class, labels=c("Not Depressed", "Depressed")), 
                    truth = factor(truth, labels=c("Not Depressed", "Depressed")))

table(plot.mpp$pred.class, plot.mpp$truth)
```

The model correctly identified 195 individuals as not depressed and 15 as depressed. The model got it wrong 49 + 35 times.

## Accuracy {.smaller}

The **accuracy** of the model is calculated as the fraction of times the model prediction matches the observed category:

```{r}
(195+15)/(195+35+49+15)
```

This model has a 71.4% accuracy.

* Is this good? 
* What if death were the event?

We will see later how to use accuracy as a measure of model fit and how to optimize accuracy by changing the cutoff value. 