---
title: "HW 04: Univariate Graphing"
subtitle: "Let the data be beautiful"
execute:
  echo: true
  warning: false
  message: false
---

# Purpose

There are a variety of conventional ways to visualize data - tables, histograms, bar graphs, etc. The purpose is always to examine the distribution of variables related to your research question. You will create a plot, follow up each graphic with a table of summary statistics (for quantitative variables) or frequency and proportion table (for categorical), and then a summary paragraph that brings it all together. 

----

# Instructions 


Completely describe 2 categorical and 2 quantitative variables using all of the following: 

* An explanation of what the variable is, and how it is measured. 
* A table of summary statistics,
* An appropriate plot with titles and axes labels, 
* A short paragraph description in full complete English sentences. 
    
    
To guide your description of this distribution try to include the following information:

* What is the trend in the data? What exactly does the chart show? (Use the chart title to help you answer this question)
* Describe the location of the bulk of the data. Measures include
    - Measures of center: mean/median
    - IQR as Q3 and Q1 (i.e., 50% of the data lie between $x_{1}$ and $x_{2}$)
    - N and (%) for categorical data for the largest category. 
* Describe the shape (for quantitative measures only):
    - Symmetry/Skewness - Is it symmetric, skewed right, or skewed left?
    - Modality - Is it uniform, unimodal, or bimodal?       
* Describe the spread or variability in the data
    - Appropriate measures include range, standard deviation, IQR for continuous data
    - For categorical data describe if distribution is spread across multiple response categories or mainly only one.
* Describe the outliers (note: there may not be any for every graph). Continuous data only.
    - Are there any outliers for the variable?
    - If yes, are these true outliers or false (due to data management or input error) outliers? 
    (This could alert you to missing codes like -77 or 99 that need to be set to missing)

----

# Submission instructions

### Draft

* Use the template provided: [[QMD]](_templates/hw04_univ_graphing_template.qmd)
  - Right click and 'save as', put this in your `scripts` folder
  - Rename this file to `univ_graphing_userid` BEFORE you open it. 
* Upload your draft PDF to `04 Univariate Graphing/draft` folder in Google Drive.


### Peer Review instructions

As a reviewer, this is what you're checking for: 

* Did they describe 2 categorical and 2 quantitative variables? 
* Did they use appropriate plot types?
* Do the variables in the plot match their description? 
* Do they have a sentence description that reads reasonably well? 
* Did they include all the required summary numbers in their description? 


### Final

* Upload your final PDF to `04 Univariate Graphing/final` folder in Google Drive.


----

\newpage 

# Example
This example uses the `mpg` data set from the `ggplot2` package.
```{r}
# Load libraries
library(sjPlot); library(ggplot2);library(ggpubr)

# Import data
# load(here("data", "data.Rdata")) # your import line will look something like this
```

> text written like this are example student submission text

## Basic categorical
Draft style plot, direct computer output showing/copied. Poor grammar and/or sentence structure, no attempt at explaining what the variable means, extra unnecessary or incorrect information included. Typos. 

```{r}
table(mpg$class)
```

> class has more suvs than compacts. 2% are 2seaters. there are 5 2seaters 47 cmpact 41 midize 11 minivans 33 pickups 35% subcompacts, 62 suv and 234 total cars. 

## Proficient categorical
Cleaned up plot, full English sentences, useful text formatting of variable names and levels. Explained what the variable was named and what it measured. 

> The `class` variable from the `mpg` data set is a catgorical variable that describes the type of vehicle being measured. Some levels of this categorical variable include _compact_, _pickup_ and _suv_. 

```{r, fig.height=4, fig.align='center'}
plot_frq(mpg$class)
```

> Sub compact cars are the most frequently reported type of car, making up over one-quarter (26.5%) of the cars in this data set with n=62 cars represented. The least represented car is a compact car with n=5 (2.1%) records. 

\newpage

## Basic quantitative 
A "for your eyes only" quality plot, minimal description of characteristics, no information about the variable provided. 

```{r}
hist(mpg$cty)
summary(mpg$cty)
```

> cty is normally distributed between 5 and 35 with mean 17.

## Proficient quantitative
Overlaid a density curve on the histogram, added a dashed line at the mean. Table of summary statistics present in a nicely formatted way, digits rounded appropriately. Plot cleaned up with appropriate axis and titles.  

> The `cty` variable records the miles per gallon (mpg) achieved during city driving. This is a quantititative numeric variable. 

```{r, fig.height=3, fig.width=4, fig.align='center'}
# this function is from the ggpubr package
gghistogram(data=mpg, x = "cty", add = "mean", fill = "skyblue", add_density = TRUE, 
            xlab = "MPG", title = "City miles per gallon (MPG)")

summary(mpg$cty)
```

> The MPG in the city ranges from 9 to 35, unimodal and is slightly skewed right with a mean of 16.9 close to the median of 17 and a standard deviation of 4.3mpg. The histogram indicates that there are at a few upper end potential outliers achieving a city MPG of approximately over 30 mpg. 

    