---
title: "HW 05: Describing Relationships"
subtitle: "Exploring Associations between two variables"
execute:
  echo: true
  warning: false
  message: false
---

# Purpose

To fully explore the relationship between two variables both summary statistics and visualizations are important. 

----

# Instructions 

For this assignment you will describe the relationship between these **four** specific combinations of data types: 

* Categorical response and categorical explanatory variable. (C ~ C)
* Quantitative response and categorical explanatory variable. (Q ~ C)
* Any combination of the above with a binary variable (B ~ C, C ~ B, or Q ~ B)
* Quantitative response and quantitative explanatory variable. (Q ~ Q)


## Before you start, 

1. Determine what variables you want to graph based on your research topic. 
    * You will need a mixture of categorical and quantitative variables for this assignment. 
    * You should use variables that are relevant to your research topic. 
    * If you have not yet identified both a quantitative (Q), a binary (B), and a categorical (C) variable that you are interested in, now is the time to go back to the codebook and figure this out. 
2. Recode variables as needed. 
    * If your response variable is categorical with many levels, you may want to collapse it down to fewer than 5 levels. 

If you do not already have a binary (B) variable in your clean data you can either a) go back and edit your `dm` file to include a binary varible, or b) dichotomize one of your categorical (C) varibles into two levels. (Make a new variable, don't overwrite your categorical variable).   



## For each bivariate relationship under consideration you will do the following: 

1. Name and explain the two variables under consideration. 

2. Create the appropriate graphic for bivariate relationship under consideration. For these plots binary variables are treated as categorical variables with only 2 levels. 
    * C ~ C: Side by side barplot
    * Q ~ C: Paneled histogram with density overlaid, or a grouped boxplot with overlaid violin plot. 
    * Q ~ Q: Scatterplot. Add both lowess and linear trend lines.

3. Calculate appropriate grouped summary statistics
    * For continuous outcomes you'll want to describe measures including the sample size, mean, median, range and variance for each level of the categorical variable. 
    * For categorical outcomes you'll want to calculate %'s of your outcome measurement across levels of your covariate. 
        - i.e. proportion of males who are smokers compared to proportion of females who are smokers
        - or proportion of smokers who are male, compared to proportion of non-smokers who are male. 

4. Explain the relationship or trends you see in the data in a summary paragraph. Put this paragraph below the graphic. 
    * Use summary statistics in your text explanation. 
    * Use specific features of the graphic in your text explanation.
        - i.e. are there outliers only in one group?
        - Do the data seem clumped or clustered in one region of the scatterplot? 
        - Is there a linear or non-linear pattern? 
        - Does one combination of categorical levels (C~C) seem to hold most the data? 
        - Are there any outlying data points? Don't list off each one, just state if there is and where approximately it's at. 

----

# Submission instructions


* Use the template provided: [[QMD]](_templates/hw05_describing_relationships_template.qmd)
  - Right click and 'save as', put this in your `scripts` folder
* Upload your final PDF to Canvas by the due date. 


----

## Example

In these examples I use a combination of `ggplot2` and `sjPlot` for plotting, `dplyr` for data management and to create summary statistics, `knitr` to create nice tables, `gridExtra` to put plots side by side. You do not need to report this information in your assignment. This is for your knowledge only. I also explain what I'm doing in code here for you to learn. you do not need to explain your own code. 
```{r, message=FALSE, warning=FALSE}
library(ggplot2); library(dplyr); library(knitr); library(ggdist)
library(scales); library(gridExtra); library(sjPlot); library(gtsummary)
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```


### C ~ C Association
For this example I am using the `NC Births` data set, data on 1000 births in 2004 from North Carolina. 
This example explores the association between the smoking status of the mother (`habit`) and whether or not the baby was born prematurely (`premie`). I specifically want to compare the percent of premie babies within the smoking and non-smoking groups. 

```{r, echo=FALSE}
nc <- read.csv("https://norcalbiostat.netlify.app/data/NCbirths.csv", header=TRUE)

nc %>% select(habit, premie) %>%
  tbl_summary(by = "habit")

plot_xtab(x=nc$habit, grp=nc$premie, show.total=FALSE, margin='row')
```

Contrary to what I was expecting, there is equal proportion of prematurely born babies to non-smokers (133/873, 15%) compared to babies born to smokers (19/126, 15%). There is no association between the smoking status of the mother and the likelihood of the baby being born prematurely. 



### Q ~ C Association

This example explores the association between the depth of a penguins bill and the species of the penguin. The quantitative response variable is bill depth in mm (`bill_depth_mm`) and the categorical explanatory variable is species (`species`). 

```{r, fig.width=5, fig.height=3, fig.align='center', warning=FALSE, message=FALSE}
pen <- palmerpenguins::penguins
pen %>% group_by(species) %>% 
  summarize(n=n(), 
            mean = mean(bill_depth_mm, na.rm = TRUE), 
            median = median(bill_depth_mm, na.rm = TRUE), 
            sd = sd(bill_depth_mm, na.rm = TRUE), 
            IQR = IQR(bill_depth_mm, na.rm = TRUE))

ggplot(pen, aes(x=bill_depth_mm, y=species, fill=species)) + 
      stat_slab(alpha=.5, justification = 0) + 
      geom_boxplot(width = .2,  outlier.shape = NA) + 
      geom_jitter(alpha = 0.5, height = 0.05) +
      stat_summary(fun="mean", geom="point", col="red", size=4, pch=17) + 
      theme_bw() + 
      labs(x="Bill depth (mm)", y = "Species") + 
      theme(legend.position = "none")
```

The distribution of bill depth are fairly normal for each species, with some higher end values causing a slight right skew for Adelie and Gentoo. Gentoo penguins have lower average bill depth compared to Adelie or Chinstrap (15.0mm vs 18.3 and 18.4mm respectively). Chinstrap however have a larger IQR at 1.9 compared to 1.5 for the others. 

### Q ~ Q Association

This example explores the association between the length of a penguins flipper and it's body mass. The quantitative response variable is body mass(`body_mass_g`) and the quantitative explanatory variable is flipper length (`flipper_length_mm`).

```{r}
cor(pen$flipper_length_mm, pen$body_mass_g, 
    use="pairwise.complete.obs") # calculate the correlation

ggplot(pen, aes(x=flipper_length_mm, y=body_mass_g)) + geom_point() + 
  geom_smooth(se=FALSE, col="brown") + geom_smooth(se=FALSE, method="lm") +
  theme_bw()
```

There is a positive association between flipper length and body mass of a penguin. The correlation coefficient is 0.87, and the form of the relationship is relatively linear.




    