---
title: "Study Design" # title 
subtitle: "How did the data come to be? " # sub title
description: "10/03" #date range
---


```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse);library(readxl); 
library(pander); library(janitor)

# Function to clean strings & separate into lists
str_prep <- function(var){
  as.list(
    str_split(var, "\r\n", simplify=TRUE)
    )
}

topic_mats <- read_excel("../schedule_615.xlsx", sheet="weekly_schedule") %>%
  clean_names() %>% 
  filter(grepl("t06", topic)) %>%
  mutate(assignment = gsub("hw", "../hw/hw", hw), 
         slides = sprintf("[%s](%s)", lecture, paste0("../slides/", lecture, ".html")), 
         collab = sprintf("[%s](%s)", collab_notes, collab_notes)
         ) %>% 
  select(slides, collab, reading, assignment, topic_obj)
```

## Learning Path

### Where we've been
* Learning how to explore and visualize our data. 
* Visually consider patterns, identify potential relationships

### Where we're at
* Think about how data come to be. That is, if the data are to be used to make broad and complete conclusions, then it is important to understand who or what the data represent. 


### Objectives

```{r}
pander(str_prep(topic_mats$topic_obj))
```

## Learning Materials


:books:  **Reading** 

```{r}
pander(str_prep(topic_mats$reading))
```


:pencil: **Collaborative notes**

```{r}
pander(str_prep(topic_mats$collab))
```

## Assignments

None. All work is done in the collab notes. 

## Discussion

