---
title: "Moderation and Stratification" # title 
subtitle: "Identifing different trends within subgroups" # sub title
description: "10/30 - 11/1" #date range
---

```{r}
source(here::here("topics/common_code.R"))
this_topic <- topic_mats %>% filter(grepl("t08", topic))
```

## Learning Path

### Where we've been
Learning how to choose and implement an appropriate bivariate analyses, including building simple regression models.  

### Where we're at
Ready to learn how to expand regression models to include information contained in other variables. 



### Objectives

```{r}
create_tabled_list(this_topic$topic_obj)
```


## Learning Materials

{{< fa person-chalkboard >}}  **Slides (Will open in full screen. Right click to open in a new tab)**

```{r}
create_tabled_list(this_topic$slides)
```

:books:  **Reading** 

```{r}
create_tabled_list(this_topic$reading)
```


## Assignments

```{r}
create_tabled_list(this_topic$assignment)
```

## Assessment

```{r}
create_tabled_list(this_topic$quiz)
```
