---
title: "Model Building" # title 
subtitle: "Part science, part art" # sub title
---

```{r}
source(here::here("topics/common_code.R"))
this_topic <- topic_mats %>% filter(grepl("t11", topic))
```

## Learning Path

### Where we've been  
Learning how to fit multivariable models for a variety of outcome types. 

### Where we're at
Learning how to build a "good" model, and how to choose between models. 

### Objectives

```{r}
create_tabled_list(this_topic$topic_obj)
```


## Learning Materials
Directly from [ASCN](https://norcalbiostat.github.io/AppliedStatistics_notes/) Chapter 10


### Model Building

* More on outliers and influential points: PMA6 Ch 7.8 and 8.8
* Multicollinearity (highly correlated X values): PMA6 8.9 
* How to choose variables for your model sensibly: PMA6 Chapter 9

### Measures of Model Fit

* [Linear Regression](https://norcalbiostat.github.io/AppliedStatistics_notes/model-fit-criteria.html) for low RMSE, large $R^{2}$, small AIC and BIC.  
* Logistic Regression: [Goodness of fit](https://norcalbiostat.github.io/AppliedStatistics_notes/logreg.html), and [Accuracy](https://norcalbiostat.github.io/AppliedStatistics_notes/predicted-class-outcome.html)


:books:  **Reading** 

```{r}
create_tabled_list(this_topic$reading)
```

## Assignments


```{r}
create_tabled_list(this_topic$assignment)
```
