---
title: "Model Building" # title 
subtitle: "Part science, part art" # sub title
description: "11/28 - 11/30" #date range
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse);library(readxl); 
library(pander); library(janitor)

# Function to clean strings & separate into lists
str_prep <- function(var){
  as.list(
    str_split(var, "\r\n", simplify=TRUE)
    )
}

topic_mats <- read_excel("../schedule_615.xlsx", sheet="weekly_schedule") %>%
  clean_names() %>% 
  filter(grepl("t12", topic)) %>%
  mutate(assignment = gsub("hw", "../hw/hw", hw), 
         slides = sprintf("[%s](%s)", lecture, paste0("../slides/", lecture, ".html")), 
         collab = sprintf("[%s](%s)", collab_notes, collab_notes)
         ) %>% 
  select(slides, collab, reading, assignment, topic_obj)
```

## Learning Path

### Where we've been
* Practicing building multivariable regression models

### Where we're at
Ready to learn how to refine our regression models to 


### Objectives

```{r}
pander(topic_mats$topic_obj)
```


## Learning Materials

:books:  **Reading** 

```{r}
pander(str_prep(topic_mats$reading))
```

### Model Building

* More on outliers and influential points: PMA6 Ch 7.8 and 8.8
* Got a moderator? Check out Interaction terms in PMA6 Ch 8.8 and ASCN 10.3
* Multicollinearity (highly correlated X values): PMA6 8.9 and ASCN 10.5
* How to choose variables for your model sensibly: PMA6 Chapter 9, ASCN 10.4, 10.6

### Measures of Model Fit

* Linear Regression ASCN 10.7
* Logistic Regression ASCN 11.3.7 for goodness of fit, and 12.1-12.2 for accuracy

