---
title: "Preparing Data for Analysis" # title 
subtitle: "Where an inordinate amount of time is spent" # sub title
---

```{r}
library(pander)
source("common_code.R")
this_topic <- topic_mats %>% filter(grepl("t03", topic))
```

## Learning Path

### Where we've been
Learning how to read our data and ask questions that are potentially answerable with this data. 

### Where we're at
* Starting to write code.
* Recognizing that this stage will take time and that you are learning something challenging.
* Willing to ask for help and collaboration


---

### Objectives

```{r}
create_tabled_list(this_topic$topic_obj)
```

## Preparation 

In Homework 00 you downloaded and installed R and R Studio, and used it to import a data file in Homework 01. There are a few more steps to take that will allow you to be best setup for using R to analyze data.  

Follow the instructions in the Appendix of the [Applied Stats Course Notes](https://norcalbiostat.github.io/AppliedStatistics_notes/setup_intro.html) to do the following: 

1. Set preferences for sanity (19.5)
2. Install the `tidyverse` and `here` packages (19.6)
3. Create a R project using the `MATH615` folder that you created in step 1. (ASCN 19.7)
   - Shut down R studio fully, navigate to your class folder, and open your project file (the cube) before you continue. 
4. Read through and follow the ["Hello Quarto" tutorial](https://quarto.org/docs/get-started/hello/rstudio) (19.8 intro)
7. Install the `tinytex` program so you can [create PDFs](https://norcalbiostat.github.io/AppliedStatistics_notes/literate-programming-with-quarto.html#creating-pdfs) from your Quarto document. This is required (19.8)

If you have difficulty with any of this, visit Community Coding, my office hours, or post in Discord. 

## Learning Materials

{{< fa person-chalkboard >}}  **Slides (Will open in full screen. Right click to open in a new tab)**

* [lec03a-workflow](../slides/lec03a-workflow.html)  
* [lec03b-data_prep](../slides/lec03b-data_prep.html)


We will also be using the Data Management chapter of the [Applied Stats notebook](https://norcalbiostat.github.io/AppliedStatistics_notes/data-management.html) for this topic.

:books:  **Reading** 

```{r}
create_tabled_list(this_topic$reading)
```


:pencil: **Collaborative notes**

```{r}
create_tabled_list(this_topic$collab)
```

## Assignments

```{r}
create_tabled_list(this_topic$assignment)
```

## Assessment
This quiz will contain questions that reference topics in the PMA6. Don't rely only on the slides for your answers. 

```{r}
create_tabled_list(this_topic$quiz)
```

## Other references

* ANYTHING by [Danielle Navarro](https://slides.djnavarro.net/) or [Jenny Bryan](https://www.rstudio.com/authors/jenny-bryan/)
* [R projects](https://martinctc.github.io/blog/rstudio-projects-and-working-directories-a-beginner's-guide/)
* [Why should I use the here package when I'm already using projects?](https://malco.io/2018/11/05/why-should-i-use-the-here-package-when-i-m-already-using-projects/)
* [Hello, Quarto](https://quarto.org/docs/get-started/hello/rstudio.html)



### Where you'll start

<iframe width="560" height="315" src="https://www.youtube.com/embed/m_MaJDK3VNE" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

### Where you'll end up

<iframe width="560" height="315" src="https://www.youtube.com/embed/3Iz7ZMALaCY" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>


