---
title: "Data Management" # title 
subtitle: "Where an inordinate amount of time is spent" # sub title
description: "Aug 31 - Sep 11" #date range
---


```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse);library(readxl); 
library(pander); library(janitor)

# Function to clean strings & separate into lists
str_prep <- function(var){
  as.list(
    str_split(var, "\r\n", simplify=TRUE)
    )
}

topic_mats <- read_excel("../schedule_615_test.xlsx", sheet="weekly_schedule") %>%
  clean_names() %>% 
  filter(grepl("t03", topic)) %>%
  mutate(assignment = gsub("hw", "../hw/hw", hw), 
         slides = sprintf("[%s](%s)", topic_name, paste0("../slides/", lecture, ".html"))) %>% 
  select(slides, collab, reading, assignment, topic_obj)
```

## Learning Path

### Where we've been
Learning how to read our data and ask questions that are potentially answerable with this data. 

### Where we're at
* Starting to write code.
* Recognizing that this stage will take time and that you are learning something challenging.
* Tapping into our grit, resolve and patience. 
* Willing to ask for help and collaboration


:::{.callout-important icon=false}
## Monday 9/5 is Labor Day. No Class. 
Enjoy the three day weekend, and consider learning something new about this holiday. (Optional reading)

* [US Department of Labor: "History of Labor Day"](https://www.dol.gov/general/laborday/history)
* [America's Black Holocaust Museum: "America's Racist History of Labor"](https://www.abhmuseum.org/watch-americas-racist-history-of-labor/)
* [Smithsonian: "Striking Union Workers Turned the First Labor Day Into a Networking Event "](https://www.smithsonianmag.com/history/striking-union-workers-turned-first-labor-day-networking-event-180964705/)
* [UAW: "Celebrating Labor day by Celebrating Labor Unions"](https://uaw.org/celbrating-labor-day-celebrating-labor-unions/)

Got more? Share in our Discord channel.
:::


---

### Objectives

```{r}
pander(str_prep(topic_mats$topic_obj))
```


## Learning Materials

{{< fa person-chalkboard >}}  **Slides (Will open in full screen. Right click to open in a new tab)**

```{r}
pander(str_prep(topic_mats$slides))
```

We will also be using the Data Management chapter of the [Applied Stats notebook](https://norcalbiostat.github.io/AppliedStatistics_notes/data-management.html) for this topic.

:books:  **Reading** 

```{r}
pander(str_prep(topic_mats$reading))
```


:pencil: **Collaborative notes**

```{r}
pander(str_prep(topic_mats$collab))
```

## Assignments

```{r}
pander(str_prep(topic_mats$assignment))
```

## Discussion

* Do not underestimate the importance of this step
* It will take you far, far longer than you anticipate to 'clean' your data
* Writing code (in any language) will be challenging, but will pay off in the long run

### It may feel like this now

<iframe width="560" height="315" src="https://www.youtube.com/embed/m_MaJDK3VNE" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

### But it'll feel like this later

<iframe width="560" height="315" src="https://www.youtube.com/embed/3Iz7ZMALaCY" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>


