---
title: "Describing distributions of data" # title 
subtitle: "Visualizing your data is the first line of defense against bad data" # sub title
---

```{r}
library(pander)
source(here::here("topics/common_code.R"))
this_topic <- topic_mats %>% 
  filter(Order %/% 1 == 4, 
         !grepl("PROJECT", Topic))

header <- this_topic[1,]
content <- this_topic[-1,]
```

## Learning Path

### Where we've been
Importing & starting to wrangle our data

### Where we're at
Ready to visualize and learn characteristics of our data. 


### Objectives

```{r}
create_tabled_list(header$topic_obj)
```

## Preparation 

A basic statistics class is a pre-requisite for this class. Ideally this was within the last 4-5 years, but I recognize not all learning paths are ideal. 
If you are rusty on descriptive statistics such as mean, median, skewness, standard deviation and variance, how to read boxplots, histograms and the IQR rule then I highly recommend that you review [IMS Chapter 5](https://openintro-ims.netlify.app/explore-numerical). 

The quiz for this topic includes topics that are not directly lectured on but are considered critical knowledge and are included in either PMA6 Ch 4 or IMS Ch 5. 


## Learning Materials

{{< fa person-chalkboard >}}  **Slides (Will open in full screen. Right click to open in a new tab)**

```{r}
create_tabled_list(content$Slides)
```

:books:  **Reading** 

```{r}
create_tabled_list(content$Reading)
```


## :pencil: Active Learning

```{r}
create_tabled_list(content$Activities)
```

## Assessment

```{r}
create_tabled_list(content$Assess)
```

