---
title: "Describing Relationships" # title 
subtitle: "First step in investigating a question about an association" # sub title
---

```{r}
library(pander)
source(here::here("topics/common_code.R"))
this_topic <- topic_mats %>% 
  filter(Order %/% 1 == 5, 
         !grepl("PROJECT", Topic))

header <- this_topic[1,]
content <- this_topic[-1,]
```

## Learning Path

### Where we've been
* Visualizing a single variable
* Screening for potential errors and additional data management
* Starting to describe our project data

### Where we're at
* Ready to visualize and describe associations between multiple variables. 


### Objectives

```{r}
create_tabled_list(header$topic_obj)
```


## Learning Materials

{{< fa person-chalkboard >}}  **Slides (Will open in full screen. Right click to open in a new tab)**

```{r}
create_tabled_list(content$Slides)
```

:books:  **Reading** 

```{r}
create_tabled_list(content$Reading)
```


## :pencil: Active Learning

```{r}
create_tabled_list(content$Activities)
```

## Assessment

```{r}
create_tabled_list(content$Assess)
```

