---
title: "Study Design" # title 
subtitle: "How did the data come to be?" # sub title
description: "9/25-9/27" #date range
execute: 
  echo: false
  message: false
  warning: false
---


```{r}
source("common_code.R")
this_topic <- topic_mats %>% filter(grepl("t06", topic))
```

## Learning Path

### Where we've been
* Learning how to explore and visualize our data. 
* Visually consider patterns, identify potential relationships

### Where we're at
* Think about how data come to be. That is, if the data are to be used to make broad and complete conclusions, then it is important to understand who or what the data represent. 
    * Knowing how the observational units were selected will allow for generalizations back to the population from which the data were selected. 
    * By understanding the structure of the study, causal relationships can be separated from those relationships which are only associated. 


### Objectives

```{r}
create_tabled_list(this_topic$topic_obj)
```

## Learning Materials


:books:  **Reading** 

```{r}
create_tabled_list(this_topic$reading)
```


:pencil: **Collaborative notes**

We will be using the [Jigsaw](https://en.wikipedia.org/wiki/Jigsaw_(teaching_technique)) technique, which is a method of [learning by teaching](https://en.wikipedia.org/wiki/Learning_by_teaching). You have been assigned (possibly with a partner) to read and summarize one section in the textbook. 

```{r}
create_tabled_list(this_topic$collab)
```

## Assignments

```{r}
create_tabled_list(this_topic$assignment)
```

## Assessment

```{r}
create_tabled_list(this_topic$quiz)
```

