---
title: "Foundations for Statistical Inference" # title 
subtitle: "Study Design, Inference using randomization and the Normal Model" # sub title
description: "9/25-9/27" #date range
---


```{r}
source("common_code.R")
this_topic <- topic_mats %>% filter(grepl("t06", topic))
```

## Learning Path

### Where we've been
* Learning how to explore and visualize our data. 
* Visually consider patterns, identify potential relationships

### Where we're at
* Think about how data come to be. That is, if the data are to be used to make broad and complete conclusions, then it is important to understand who or what the data represent. 
    * Knowing how the observational units were selected will allow for generalizations back to the population from which the data were selected. 
    * By understanding the structure of the study, causal relationships can be separated from those relationships which are only associated. 
* Revisiting null value hypothesis testing (NVHT) from a randomization perspective
* Revisiting the Normal distribution model, and getting a feel for how probability (and thus p-values) are calculated. 



### Objectives

```{r}
create_tabled_list(this_topic$topic_obj)
```

## Learning Materials


:books:  **Reading** 

```{r}
create_tabled_list(this_topic$reading)
```


:cinema: **Video**

* [What is a sampling distribution](https://youtu.be/xGVzHxFnaiA)
* [Sampling Distributions: Introduction to the concept](https://youtu.be/Zbw-YvELsaM)


:pencil: **Collaborative notes**

We will be using the [Jigsaw](https://en.wikipedia.org/wiki/Jigsaw_(teaching_technique)) technique, which is a method of [learning by teaching](https://en.wikipedia.org/wiki/Learning_by_teaching). You have been assigned (possibly with a partner) to read and summarize one section in the textbook. 

```{r}
create_tabled_list(this_topic$collab)
```

## Assignments

```{r}
create_tabled_list(this_topic$assignment)
```

## Assessment

```{r}
create_tabled_list(this_topic$quiz)
```




