---
title: "Foundations for Inference" # title 
subtitle: "[If all else fails, use 'significant at p>.05 level' and hope no one notices.](https://xkcd.com/1478/)" # sub title
description: "10/2-10/4" #date range
execute: 
  echo: false
  message: false
  warning: false
---


```{r}
source("common_code.R")
this_topic <- topic_mats %>% filter(grepl("t06b", topic))
```

## Learning Path

### Where we've been
* Thinking about how Statistics uses data from small random samples to make conclusions about a larger population of interest. 
* Understanding the limitations of observational studies, and the difference between correlation and causation. 

### Where we're at
* Revisiting null value hypothesis testing (NVHT) from a randomization perspective
* Revisiting the Normal distribution model, and getting a feel for how probability (and thus p-values) are calculated. 



### Objectives

```{r}
create_tabled_list(this_topic$topic_obj)
```

## Learning Materials

:books:  **Reading** 

```{r}
create_tabled_list(this_topic$reading)
```

:cinema: **Video**

* [What is a sampling distribution](https://youtu.be/xGVzHxFnaiA)
* [Sampling Distributions: Introduction to the concept](https://youtu.be/Zbw-YvELsaM)


:pencil: **Collaborative notes**


```{r}
create_tabled_list(this_topic$collab)
```

## Assignments

Continuation of HW06

## Assessment

```{r}
create_tabled_list(this_topic$quiz)
```

