---
title: "Modeling Bivariate Relationships" # title 
subtitle: "Identify and conduct the most appropriate analysis for a given research topic. " # sub title
description: "10/11 - 10/25" #date range
---

```{r}
source("common_code.R")
this_topic <- topic_mats %>% filter(grepl("t07", topic))
```

## Learning Path

### Where we've been
* Visually exploring relationships between two variables by creating appropriate plots to assess how the distribution of a primary outcome (response/dependent) variable changes according to the level of a predictor (explanatory/independent/covariate) variable.
* Learning how probability is used to make inferential statements about parameters

### Where we're at
Ready to make some statistical statements about relationships between variables. 


### Objectives

```{r}
create_tabled_list(this_topic$topic_obj)
```


## Learning Materials

{{< fa person-chalkboard >}}  **Slides (Will open in full screen. Right click to open in a new tab)**

```{r}
create_tabled_list(this_topic$slides)
```

:books:  **Reading** 

```{r}
create_tabled_list(this_topic$reading)
```

:cinema: **Video**

Pre-recorded lectures from last year are available in Canvas. The slides are mostly the same, but this is dense content and so these videos give you a chance to review. 


:pencil: **Collaborative notes**

```{r}
create_tabled_list(this_topic$collab)
```

## Assignments

```{r}
create_tabled_list(this_topic$assignment)
```

## Assessment

```{r}
create_tabled_list(this_topic$quiz)
```

