---
title: "Inference with Mathematical Models" # title 
subtitle: "If all else fails, use 'significant at p>.05 level' and hope no one notices. Ref: https://xkcd.com/1478/?" # sub title
description: "10/05-10/10" #date range
---


```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse);library(readxl); 
library(pander); library(janitor)

# Function to clean strings & separate into lists
str_prep <- function(var){
  as.list(
    str_split(var, "\r\n", simplify=TRUE)
    )
}

topic_mats <- read_excel("../schedule_615.xlsx", sheet="weekly_schedule") %>%
  clean_names() %>% 
  filter(grepl("t07", topic)) %>%
  mutate(assignment = gsub("hw", "../hw/hw", hw), 
         slides = sprintf("[%s](%s)", topic_name, paste0("../slides/", lecture, ".html"))) %>% 
  select(slides, collab, reading, assignment, topic_obj, video)
```

## Learning Path

### Where we've been
Learning about different data generating mechanisms/models. 

### Where we're at
Refreshing our memory from a prior stat class how to use the Normal distribution to compare values, calculate probabilities, confidence intervals and conduct a hypothesis test 

### Objectives

```{r}
pander(str_prep(topic_mats$topic_obj))
```

## Learning Materials


:books:  **Reading** 

```{r}
pander(str_prep(topic_mats$reading))
```

:pencil: **Collaborative notes**

```{r}
pander(str_prep(topic_mats$collab))
```

## Assignments

None. All work is done in the collab notes. 

::: {.callout-important}
## Quizzes

To assess your understanding of this topic (and not take up class time) here are two required quizzes. You have until the date on the calendar to take these quizzes. You can repeat them as many times as you like until you get the grade you are happy with. 

Note: Some questions require manual review (because they are written answers). I will check nightly for new responses and do the manual grading so you can see your score. 

Quantifying Uncertainty: https://forms.gle/PQGYizu45bCSoQAM9
Foundations for Inference: https://forms.gle/a4tEWsRf1nxkoz3P6 
:::

```{r}
pander(str_prep(topic_mats$assignment))
```


## Discussion

Additional videos that may be helpful. 

* [What is a sampling distribution](https://youtu.be/xGVzHxFnaiA)
* [Sampling Distributions: Introduction to the concept](https://youtu.be/Zbw-YvELsaM)
