---
title: "Inference with Mathematical Models" # title 
subtitle: "If all else fails, use 'significant at p>.05 level' and hope no one notices. Ref: https://xkcd.com/1478/?" # sub title
description: "10/05-10/10" #date range
---


```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse);library(readxl); 
library(pander); library(janitor)

# Function to clean strings & separate into lists
str_prep <- function(var){
  as.list(
    str_split(var, "\r\n", simplify=TRUE)
    )
}

topic_mats <- read_excel("../schedule_615.xlsx", sheet="weekly_schedule") %>%
  clean_names() %>% 
  filter(grepl("t07", topic)) %>%
  mutate(assignment = gsub("hw", "../hw/hw", hw), 
         slides = sprintf("[%s](%s)", topic_name, paste0("../slides/", lecture, ".html"))) %>% 
  select(slides, collab, reading, assignment, topic_obj, video)
```

## Learning Path

### Where we've been
Learning about different data generating mechanisms/models. 

### Where we're at
Refreshing our memory from a prior stat class how to use the Normal distribution to compare values, calculate probabilities, confidence intervals and conduct a hypothesis test 

### Objectives

```{r}
pander(str_prep(topic_mats$topic_obj))
```

## Learning Materials


:books:  **Reading** 

```{r}
pander(str_prep(topic_mats$reading))
```

:pencil: **Collaborative notes**

```{r}
pander(str_prep(topic_mats$collab))
```

## Assignments

None. All work is done in the collab notes. 

## Discussion

Additional videos that may be helpful. 

* [What is a sampling distribution](https://youtu.be/xGVzHxFnaiA)
* [Sampling Distributions: Introduction to the concept](https://youtu.be/Zbw-YvELsaM)
