---
title: "Modeling Bivariate Relationships" # title 
subtitle: "Identify and conduct the most appropriate analysis for a given research topic. " # sub title
---

```{r}
source(here::here("topics/common_code.R"))
this_topic <- topic_mats %>% 
  filter(Order %/% 1 == 7, 
         !grepl("PROJECT", Topic))

header <- this_topic[1,]
content <- this_topic[-1,]
```

## Learning Path

### Where we've been
* Visually exploring relationships between two variables by creating appropriate plots to assess how the distribution of a primary outcome (response/dependent) variable changes according to the level of a predictor (explanatory/independent/covariate) variable.
* Learning how probability is used to make inferential statements about parameters

### Where we're at
Ready to make some statistical statements about relationships between variables. 


### Objectives

```{r}
create_tabled_list(header$topic_obj)
```



## Learning Materials

{{< fa person-chalkboard >}}  **Slides (Will open in full screen. Right click to open in a new tab)**

```{r}
create_tabled_list(content$Slides)
```

:books:  **Reading** 

```{r}
create_tabled_list(content$Reading)
```


:cinema: **Video**

Pre-recorded lectures are available in Canvas. Some recordings are from a prior semester but the slides are mostly the same. 


## Assignments

```{r}
create_tabled_list(content$Activities)
```

## Assessment

```{r}
create_tabled_list(content$Assess)
```


