---
title: "Introduction to Regression modeling" # title 
subtitle: "Everything is a linear model" # sub title
---

```{r}
source(here::here("topics/common_code.R"))
this_topic <- topic_mats %>% 
  filter(Order %/% 1 == 8, 
         !grepl("PROJECT", Topic))

header <- this_topic[1,]
content <- this_topic[-1,]
```

## Learning Path

### Where we've been  
Learning how to statistically assess the relationship between two variables. 


### Where we're at
Learning the fundamentals of linear regression, a foundational method of modeling for many types of analyses. We'll start with simple linear regression, a model that describes the relationship between two quantitative variables as a straight line.

Then we'll expand that model to include multiple predictors of varying types. 

And then we'll make it even more generalizable by transforming the response variable $y$ and modeling a log transformed outcome, and a binary outcome using the same regression modeling framework. 


### Objectives

```{r}
create_tabled_list(header$topic_obj)
```


## Learning Materials

This section uses the the [Lung function](https://www.norcalbiostat.com/data/#Lung_Function) dataset and the following packages:

* Plotting: [ggplot2](https://ggplot2.tidyverse.org), [ggdist](https://mjskay.github.io/ggdist/index.html), [sjPlot](https://strengejacke.github.io/sjPlot/),[gridExtra](https://cran.r-project.org/web/packages/gridExtra/index.html)
* Presenting results: [broom](https://broom.tidymodels.org), [gtsummary](https://www.danieldsjoberg.com/gtsummary/)
* Assumption checking: [performance](https://easystats.github.io/performance/)

See PMA6 Appendix for more details about the Lung data. 

> Note: Regression is such a big deal, there are MANY functions in MANY packages that help you get out the relevant information out of the model, and display it in many different ways. I will show several in the slides. 


{{< fa person-chalkboard >}}  **Slides (Will open in full screen. Right click to open in a new tab)**

```{r}
create_tabled_list(content$Slides)
```

:books:  **Reading** 

```{r}
create_tabled_list(content$Reading)
```


## Assignments

```{r}
create_tabled_list(content$Activities)
```

## Assessment

```{r}
create_tabled_list(content$Assess)
```
