---
title: "Multiple & Generalized Regression" # title 
subtitle: "Because life isn't bivariate" # sub title
---

```{r}
source(here::here("topics/common_code.R"))
this_topic <- topic_mats %>% 
  filter(Order %/% 1 == 9, 
         !grepl("PROJECT", Topic))

header <- this_topic[1,]
content <- this_topic[-1,]
```

## Learning Path

### Where we've been  
Learning how a third variable can modify the relationship between two explanatory and response variables. 


### Where we're at
Learning how to expand regression models to include more than 1 predictor, and how to fit other non-linear regression models. 



### Objectives

```{r}
create_tabled_list(header$topic_obj)
```


## Learning Materials


{{< fa person-chalkboard >}}  **Slides (Will open in full screen. Right click to open in a new tab)**

```{r}
create_tabled_list(content$Slides)
```

:books:  **Reading** 

```{r}
create_tabled_list(content$Reading)
```


## Assignments

```{r}
create_tabled_list(content$Activities)
```

## Assessment

```{r}
create_tabled_list(content$Assess)
```
