---
title: "Multiple Regression" # title 
subtitle: "Because life isn't bivariate" # sub title
description: "11/07 - 11/18" #date range
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse);library(readxl); 
library(pander); library(janitor)

# Function to clean strings & separate into lists
str_prep <- function(var){
  as.list(
    str_split(var, "\r\n", simplify=TRUE)
    )
}

topic_mats <- read_excel("../schedule_615.xlsx", sheet="weekly_schedule") %>%
  clean_names() %>% 
  filter(grepl("t11", topic)) %>%
  mutate(assignment = gsub("hw", "../hw/hw", hw), 
         slides = sprintf("[%s](%s)", lecture, paste0("../slides/", lecture, ".html")), 
         collab = sprintf("[%s](%s)", collab_notes, collab_notes)
         ) %>% 
  select(slides, collab, reading, assignment, topic_obj)
```

## Learning Path

### Where we've been
* Learning how to choose and implement an appropriate bivariate analyses, including building simple regression models.  
* Learning how a bivariate relationship can be altered or changed by the presence of a third variable. 

### Where we're at
Ready to learn how to expand regression models to include information contained in other variables. 


### Objectives

```{r}
pander(topic_mats$topic_obj)
```


## Learning Materials

:books:  **Reading** 

```{r}
pander(str_prep(topic_mats$reading))
```

## Assignment


```{r}
pander(str_prep(topic_mats$assignment))
```


## Discussion

